---
title: "HEART FAILURE CLINICAL  _PROJECT"
author: "ANWAR ABU ALHUSSEIN"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning=FALSE, include=TRUE)
```

## Required packages
###########################################################################
###########################################################################
###                                                                     ###
###                    HEART FAILURE CLINICAL                           ###
###                   PROJECT LINEAR REGRESSION                         ###
###                                                                     ###
###########################################################################
###########################################################################
###*                        Loading Packages
###*                        ----------------
```{r echo=TRUE, message=FALSE, warning=FALSE, include=TRUE}
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(caret))
suppressPackageStartupMessages(library(ggcorrplot))
suppressPackageStartupMessages(library(Metrics))
suppressPackageStartupMessages(library(corrplot))
suppressPackageStartupMessages(library(bannerCommenter))
suppressPackageStartupMessages(library(psych))
suppressPackageStartupMessages(library(WVPlots))
suppressPackageStartupMessages(library(PerformanceAnalytics))
suppressPackageStartupMessages(library(car))
suppressPackageStartupMessages(library(dplyr))
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(outliers))
suppressPackageStartupMessages(library(broom))
suppressPackageStartupMessages(library(equatiomatic))
```




##===============================================================
##                     Reading in the data                     ==
##===============================================================
```{r echo=TRUE, message=FALSE, warning=FALSE, include=TRUE}
download.file("https://raw.githubusercontent.com/ABUALHUSSEIN/test/main/data/heart.csv",
              destfile = "heart.csv")
```


##source(destfile)
##ls()
#======================================
```{r echo=TRUE, message=FALSE, warning=FALSE, include=TRUE}
setwd("C:/Users/WAFA/Desktop")
heart <- read.csv("heart.csv",header=TRUE)
View(heart)
```

#txt <- "Explore the data set"
#banner(txt, centre = TRUE, bandChar = "-")
##----------------------------------------------------------------
##                     Explore the data set                     --
##----------------------------------------------------------------
#txt <- "Data contents"
#banner(txt, centre = TRUE, bandChar = "=")
##===============================================================
##                        Data contents                        ==
##===============================================================
```{r echo=TRUE, message=FALSE, warning=FALSE, include=TRUE}
Hmisc::contents(heart)

glimpse(heart)

head(heart, 10)

tail(heart)

car::brief(heart)
str(heart)
```


# We have 299 obs. of  13 variables.
##===============================================================
##                       Data  Description                    ==
##===============================================================
 
## **age**: age of the patient (years)
## **anaemia**: decrease of red blood cells or hemoglobin (boolean)
## **high blood pressure**: if the patient has hypertension (boolean) creatinine phosphokinase (CPK): level of the CPK enzyme in the blood (mcg/L)
## **diabetes**: if the patient has diabetes (boolean)
## **ejection fraction**: percentage of blood leaving the heart at each contraction (percentage)
## **platelets**: platelets in the blood (kiloplatelets/mL)
## ** sex**: woman or man (binary)
## **serum creatinine**: level of serum creatinine in the blood (mg/dL)
## **serum sodium**: level of serum sodium in the blood (mEq/L)
## ** smoking**: if the patient smokes or not (boolean)
## **time**: follow-up period (days)
## **death event**: if the patient deceased during the follow-up period (boolean)


##================================================================
##                EDA: Exploratory Data Analysis                ==
##================================================================
```{r echo=TRUE, message=FALSE, warning=FALSE, include=TRUE}
heart_data <- heart[, 7:ncol(heart)]
```



# Check the column names again, 
```{r echo=TRUE, message=FALSE, warning=FALSE, include=TRUE}
colnames(heart_data)
```
#txt <- "Get the summary statistics of the variables"
#banner(txt, centre = TRUE, bandChar = "*")
##***************************************************************
##         Get the summary statistics of the variables         **
##***************************************************************

```{r echo=TRUE, message=FALSE, warning=FALSE, include=TRUE}
summary(heart_data)
Hmisc::describe(heart_data)
psych::describe(heart_data, skew = TRUE, 
                IQR = TRUE)
```




#txt <- "The Dependent Variables in this study is ejection_fraction"
#banner(txt, centre = TRUE, bandChar = ":")

##::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::
##  The Dependent Variables in this study is ejection_fraction  ::
##::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::

# We are going to explore the distribution of this variable
```{r echo=TRUE, message=FALSE, warning=FALSE, include=TRUE}
ggplot(heart_data, aes(x = ejection_fraction)) + 
  geom_histogram(aes(y = ..density..), 
                 fill = "steelblue") +
  geom_density(color = "red", lwd = 1.2) + 
  stat_function(fun = dnorm, args = list(mean = mean(heart_data$ejection_fraction), 
                                         sd = sd(heart_data$ejection_fraction)),
                color = "orange", lwd = 1.2) +
  theme_linedraw()

```


#################################################################
##                        Joint Graphs                         ##
#################################################################

```{r echo=TRUE, message=FALSE, warning=FALSE, include=TRUE}
library(WVPlots)
```

#ejection_fraction VS age
```{r echo=TRUE, message=FALSE, warning=FALSE, include=TRUE}

ScatterHist(heart_data, title = "ejection_fraction VS age", 
            xvar = "age", yvar = "ejection_fraction", 
            smoothmethod = "lm")
```


## There is a very weak linear relationshipbetween the variables(ejection_fraction VS age)

#age VS serum_creatinine
```{r echo=TRUE, message=FALSE, warning=FALSE, include=TRUE}
ScatterHist(heart_data, title = "age VS serum_creatinine", 
            xvar = "serum_creatinine", yvar = "age", 
            smoothmethod = "lm")
```


# There is  a positive linear relationship between the variables(age & serum_creatinine)

# ejection_fraction VS serum_sodium
```{r echo=TRUE, message=FALSE, warning=FALSE, include=TRUE}

ScatterHist(heart_data, title = "ejection_fraction VS serum_sodium", 
            xvar = "serum_sodium", yvar = "ejection_fraction", 
            smoothmethod = "lm", hist_color = "#00ACBB", 
            point_alpha = 0.3, 
            point_color = "#FF00CC")
```

# There is  a positive linear relationship between the variables (ejection_fraction & serum_sodium)


##***************************************************************
##                    Study the correlation                    **
##***************************************************************
```{r echo=TRUE, message=FALSE, warning=FALSE, include=TRUE}
psych::lowerCor(x = heart_data)
psych::corr.test(heart_data)$p
```


# Plotting the ScatterPlotMatrix
# First look at the help, and the arguments 
# I am going to tweak the knobs a little
```{r echo=TRUE, message=FALSE, warning=FALSE, include=TRUE}
pairs(heart_data)  
```

#change the color and get halp matrix
```{r echo=TRUE, message=FALSE, warning=FALSE, include=TRUE}
pairs(heart_data, lower.panel = NULL, col= "blue")
```

# Or if you want only the lower part matrix
```{r echo=TRUE, message=FALSE, warning=FALSE, include=TRUE}
pairs(heart_data, upper.panel = NULL, col= "blue")
```

# Check the documentation for more information
```{r echo=TRUE, message=FALSE, warning=FALSE, include=TRUE}
?pairs
```

# Scatter Matrix with psych package
```{r echo=TRUE, message=FALSE, warning=FALSE, include=TRUE}
library(psych)
pairs.panels(heart_data, 
             method = "pearson", # Correlation method
             hist.col = "#11AACC",
             density = TRUE, 
             cex.cor = 1.3, 
             col = "red", 
             lm = TRUE, 
             pch = 25,    # point character
             bg = "cyan") # background
```

# Scatter Matrix with car package
```{r echo=TRUE, message=FALSE, warning=FALSE, include=TRUE}
car::scatterplotMatrix(heart_data, 
                       col = "steelblue", 
                       pch = 21, 
                       upper.panel = NULL)
```


# Lastly
```{r echo=TRUE, message=FALSE, warning=FALSE, include=TRUE}

library(PerformanceAnalytics)
chart.Correlation(heart_data, 
                  histogram=TRUE,
                  pch=19, 
                  col = "blue")

```


##***************************************************************
##                  correlation plot matrices                  **
##***************************************************************
```{r echo=TRUE, message=FALSE, warning=FALSE, include=TRUE}
correl <- cor(heart_data)

psych::cor.plot(correl)

corrplot::corrplot(correl)
```

# Tweak the knobs
```{r echo=TRUE, message=FALSE, warning=FALSE, include=TRUE}

corrplot(correl, type = "upper", 
         order = "hclust", 
         tl.col = "black",
         tl.srt = 45)
```

# check 
```{r echo=TRUE, message=FALSE, warning=FALSE, include=TRUE}
?corrplot
```

# Heatmap 
```{r echo=TRUE, message=FALSE, warning=FALSE, include=TRUE}
heatmap(correl, symm = TRUE, 
        cexRow = 0.7, 
        cexCol = 0.7)
```

# ggcorrplot
```{r echo=TRUE, message=FALSE, warning=FALSE, include=TRUE}
p <- ggcorrplot::ggcorrplot(correl, 
                            method = "circle", 
                            type = "lower", 
                            ggtheme = ggplot2::theme_linedraw, 
                            lab_col = "blue", 
                            lab_size = 3,
                            tl.cex = 10, 
                            lab = TRUE, 
                            pch.cex = 10, 
                            colors = c("#6D9EC1", "white", "#E46726")) 
p
p + guides(scale = "none")
```

##----------------------------------------------------------------
##               Fitting Multiple Regression                    --
##----------------------------------------------------------------
```{r echo=TRUE, message=FALSE, warning=FALSE, include=TRUE}
model <- lm(ejection_fraction ~ serum_sodium + age + platelets + serum_creatinine + 
              time + creatinine_phosphokinase , data = heart_data)
model
influencePlot(model)
```

##---------------------------------------------------------------
##                     Checking lm objects                     --
##---------------------------------------------------------------
# Show the components of lm object
```{r echo=TRUE, message=FALSE, warning=FALSE, include=TRUE}
str(model)
class(model)
typeof(model)
length(model)
names(model)
```

# Summary Function
```{r echo=TRUE, message=FALSE, warning=FALSE, include=TRUE}
summary(model)
```

# Coefficients Function
```{r echo=TRUE, message=FALSE, warning=FALSE, include=TRUE}
model$coefficients
```

# Fitted function
```{r echo=TRUE, message=FALSE, warning=FALSE, include=TRUE}
model$fitted
extract_eq(model,use_cof=TRUE) 
par(mfrow = c(2, 2))
plot(model)
vif(model)
```

## We will improve our result by removing the outlier
##create a boxplot to identify   outliers
```{r echo=TRUE, message=FALSE, warning=FALSE, include=TRUE}
boxplot(heart_data$ejection_fraction)
boxplot(heart_data$serum_sodium)
boxplot(heart_data$age)
boxplot(heart_data$platelets)
boxplot(heart_data$serum_creatinine)
boxplot(heart_data$time)
boxplot(heart_data$creatinine_phosphokinase)
boxplot(heart_data)
boxplot(heart_data)$out
```

###How to Remove Outliers from Multiple Columns in R

```{r echo=TRUE, message=FALSE, warning=FALSE, include=TRUE}
cooksd<-cooks.distance((model))

plot(cooksd,pch="*",cex=1,main="Influential obs by cooks distance")

cooksd
```


```{r echo=TRUE, message=FALSE, warning=FALSE, include=TRUE}
influential<-cooksd[(cooksd>(3*mean(cooksd,na.rm=TRUE)))]

names_of_influential<-names(influential)
names_of_influential
outliers <- heart_data[names_of_influential,]
heart_data_without_outliers <-heart_data %>% anti_join(outliers)
heart_data_without_outliers

```




#  Other Ways of Removing Outliers ----------------------------------------
```{r echo=TRUE, message=FALSE, warning=FALSE, include=TRUE}

outliers <- function(x) {
  
  Q1 <- quantile(x, probs=.25)
  Q3 <- quantile(x, probs=.75)
  iqr = Q3-Q1
  
  upper_limit = Q3 + (iqr*1.5)
  lower_limit = Q1 - (iqr*1.5)
  
  x > upper_limit | x < lower_limit
}

remove_outliers <- function(heart_data, cols = names(heart_data)) {
  for (col in cols) {
    heart_data <- heart_data[!outliers(heart_data[[col]]),]
  }
  heart_data
}

heart_data_without_outliers <- remove_outliers(heart_data)
boxplot(heart_data_without_outliers)
```

```{r echo=TRUE, message=FALSE, warning=FALSE, include=TRUE}
mode2 <- lm(ejection_fraction ~ serum_sodium + age + platelets + serum_creatinine + time +
              creatinine_phosphokinase , data = heart_data_without_outliers)
mode2
summary(mode2)
```



## Other Ways of Removing Outliers

## To detect the outlier,we can use the influencePlot
##influencePlot(model)

##mode2 <- lm(ejection_fraction ~ serum_sodium + age + platelets + serum_creatinine + time + creatinine_phosphokinase , data = heart_data[!(row.names(heart_data)=="218"&"65")])

##mode2

# Coefficients Function &  Fitted function
```{r echo=TRUE, message=FALSE, warning=FALSE, include=TRUE}
mode2$coefficients
mode2$fitted
extract_eq(mode2,use_cof=TRUE) 
par(mfrow = c(2, 2))
plot(mode2)
```

#### We can see here,that the coefficient is different between our new model and our previous model 
## Also, after removing the outlier from our dataset the residual standard error is decreasing, while the R_ squared is increasing
## It means that our new model fits better to our data compared to the previous model
## VIF for new model

```{r echo=TRUE, message=FALSE, warning=FALSE, include=TRUE}
vif(mode2)
glance(mode2)
tidy(mode2)
```



## vif<10 for all variables the model good.


###*************************************************************************
###*************************************************************************
###                                                                      ***
###                          SPLITTING THE DATA                          ***
###                        TRAINING AND TEST SETS                        ***
###                                                                      ***
###*************************************************************************
###*************************************************************************
###*

# Let's say we want to split the data in 80:20 for train :test dataset

```{r echo=TRUE, message=FALSE, warning=FALSE, include=TRUE}
set.seed(299)
ind <- createDataPartition(heart_data$ ejection_fraction, 
                           p = 0.8, times = 1, list = FALSE)
length(ind)
train_set <- heart_data[ind, ]
test_set <- heart_data[-ind, ]
nrow(train_set); nrow(test_set)
```


# Training the model ------------------------------------------------------
```{r echo=TRUE, message=FALSE, warning=FALSE, include=TRUE}
lm_fit <- lm(ejection_fraction ~ . , data = train_set)
broom::tidy(lm_fit)
broom::glance(lm_fit)
```



# We've built a machine learning model and trained it on train_set

###*  *** Prediction ***
#     --------------------
```{r echo=TRUE, message=FALSE, warning=FALSE, include=TRUE}
pred <- predict(object = lm_fit, newdata = test_set, type = "response") 

head(pred)
```



###       **** Model Evaluation ***
#         -------------------------
```{r echo=TRUE, message=FALSE, warning=FALSE, include=TRUE}
actual <- test_set$ejection_fraction
mae <- Metrics::mae(actual = actual, predicted = pred)
mse <- Metrics::mse(actual = actual, predicted = pred)
rmse <- Metrics::rmse(actual = actual, predicted = pred)

```


# These common metrics are used to evaluate the model.


# Table of results
```{r}
knitr::kable(cbind(mae, mse, rmse))
```




